Comparing tool variability and numerical variability in fMRI analyses

### 2021-10-12 paper/paper.tex
Comments on results

\section{Results}
All scripts and results to create the figures in this section are available at \url{https://github.com/ali4006/fuzzy-neurotools}.
The computations were performed on \href{https://www.computecanada.ca}{the Compute Canada's} Béluga cluster
with 872 available nodes containing 2× Intel Gold 6148 Skylake @ 2.4 GHz (40 cores/node) CPU and node memory ranging from 92 to 752 GB.

We ensured that fMRI analyses were replicated successfully compared to the original results in~\cite{bowring2019exploring}
by visually checking the overlap of T-statistic images resulted from the group-level analysis.
We could replicate the analyses across tools with an exception for AFNI that produced results with slight differences; 
this might be caused by the compiler variations, or different hardware and their optimizations
such as out-of-order execution (dynamic scheduling) paradigm~\cite{duben2014use,demmel2013numerical}.
In this paradigm, the processors might execute instructions out of the original order they appear based on
the availability of input data and execution units to use resources efficiently. Therefore, it might
compute floating-point operations in a different order, which often leads to different results.

Summary statistics for the group-level (un)thresholded T-statistics are reported in Table~\ref{table:pipeline-stats}.
The variability between results of each pair of tools in BT and individual tools in WT was measured using the mean and standard deviation of absolute differences.
Overall, the distribution of variations in WT was an order of magnitude lower than BT in both means and standard deviations.
While the highest variability was observed in \fslafni, \fslspm was made the least variations in BT.
Results also show that AFNI was generated the most variability in WT among the three tools.

%%%%%%%%%% Summary of statstics %%%%%%%%
\setlength{\tabcolsep}{7pt}
\begin{table}[h]
    \centering
    \begin{tabular}{cccc|cc}
        \toprule
        \multirow{2}{*}{}& {} & \multicolumn{2}{c}{Thresholded} & \multicolumn{2}{c}{Unthresholded} \\
        \cmidrule{3-4} \cmidrule{5-6} \\
        {} & {} & Mean & Std. dev. & Mean & Std. dev. \\
        \midrule
        \rowcolor{lightgray}
        {BT} & FSL vs. SPM        &  0.043       & 1.282      & 0.242     & 0.443  \\
        \rowcolor{lightgray}
        {} & FSL vs. AFNI         &  0.099       & 1.548      & 0.302     & 0.547  \\
        \rowcolor{lightgray}
        {} & AFNI vs. SPM         &  0.079       & 1.475      & 0.254     & 0.608  \\
        {WT} & FSL    &  0.005       & 0.354      & 0.034     & 0.082  \\
        {}   & SPM    &  0.006       & 0.252      & 0.025     & 0.054  \\
        {}   & AFNI   &  0.017       & 0.434      & 0.041     & 0.128  \\
        \bottomrule
    \end{tabular}
    \caption{Summary of voxel-wise mean and standard deviation of absolute differences in group-level T-statistics in BT and WT.}
    \label{table:pipeline-stats}
\end{table}

@@

%Fig 1
Comparisons of standard deviations between thresholded images in WT and BT
on MNI space are shown in Figure~\ref{fig:thresh-maps}.
While we observed substantial variations in BT with the average standard deviation $\approx$ 1.5,
the magnitude of variations in WT was much lower with the average standard deviation $\approx$ 0.5,
as was anticipated. However, numerical variability was still significant in WT results.
%Moreover, the numerical perturbations produces variations in similar order og magnitude in WT compared to BT variations
Moreover, we obtained a similar order of magnitude in WT variations compared to BT variations
in some regions in the thresholded images.
For instance, some parts of the frontal lobe in the sagittal plane for all pairs and the occipital
lobe in the axial plane for the pairs of \fslspm and \fslafni in Figure~\ref{fig:thresh-maps}
were closely replicated by the numerical perturbations.

% Fig2
In Figure~\ref{fig:dice-thresh}, we compared region-by-region Dice coefficients of
the group-level thresholded maps for all pairs of software packages in BT and WT.
This shows a linear correlation between Dice values in BT and WT,
which implies that the variability of activated voxels in both conditions was similar.
Also, small P-values across all three pairs, including \num{1e-10} for \fslafni, \num{6e-04} for \fslspm,
and \num{2e-05} for \afnispm, confirmed this correlation of similarities.
The vertical line where the Dice score is zero in BT demonstrates regions in the brain with no overlaps between activated voxels in BT but WT,
including left/right `V1\_ROI'\footnote{The primary visual cortex} in \fslafni and \fslspm,
and left/right `TGd\_ROI'\footnote{The area TG dorsal (part of lateral temporal cortex)} in \afnispm results.

%%%%%%%%%% Var. of Thresh %%%%%%%%
\begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \centering
      \includegraphics[width=\textwidth]{figures/std/gl-thresh.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \caption{Maps of standard deviation of thresholded T-statistics. The first and second rows show
    maps on BT and WT results, respectively, and the third row represents maps of their differences.}
    % so that bright blue areas indicate more similar order of magnitude of variations in both conditions,
    %and vise versa for the darkder regions.}
    \label{fig:thresh-maps}
    \end{minipage}}
  \end{figure*}

  %%%%%%%%%% Dice plot of thresholded tstats%%%%%%%%
  \begin{figure}[ht]
    \fbox{\begin{minipage}{\dimexpr \columnwidth-2\fboxsep-2\fboxrule}
    \centering
    \includegraphics[width=\columnwidth]{figures/dices_corr.png}
    \caption{Correlation of Dice coefficients of activated regions in BT and WT
    from the thresholded T-statistics. Different pairs are illustrated in different colors.
    Regions correspond to the 360 areas of cortical parcellation (HCP-MMP1.0)~\cite{glasser2016multi}.}
    \label{fig:dice-thresh}
    \end{minipage}}
  \end{figure}

@@

% Fig3
Figure~\ref{fig:unthresh-maps} shows the brain maps of standard deviations for group-level unthresholded T-statistics in WT and BT.
Results show a different order of magnitude of variations in WT and BT with standard deviation
of $\approx$ 0.12 and $\approx$ 0.5 on average, respectively.
We observed regions in the brain where the magnitude of standard deviation was close to zero in WT, but it exceeded 2.0 in BT,
such as some parts of the limbic and frontal lobes in the sagittal plane.
However, the brain maps included voxels with approximately zero differences of standard deviations in WT and BT, which are uniformly distributed.

%%%%%%%%%% Var. of Unthresh %%%%%%%%
\begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \centering
      \includegraphics[width=\textwidth]{figures/std/gl-unthresh.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \caption{Maps of standard deviation of unthresholded T-statistics. The first and second rows show
    maps on BT and WT results, respectively, and the third row represents maps of their differences.}
    \label{fig:unthresh-maps}
    \end{minipage}}
  \end{figure*}

% Fig4
The scatter plot in Figure~\ref{fig:unthresh-correlation} represents the correlation of standard deviation in WT and BT variability.
We found two major clusters, including the identity cluster that corresponds to the correlations
between BT and WT with the ratio of $0.5 < BT/WT < 2$ where the standard deviations were bigger than 0.1,
and the upper cluster that shows voxels where BT $\approx$ 0.
The percentage of voxels included in the identity cluster was \%9.9 in \fslspm, \%17.3 in \fslafni, and \%13.8 in \afnispm,
and in the upper cluster was $\approx$ \%1 for each pair of tools.
The identity area is also represented on the MNI space in the second row in Figure~\ref{fig:unthresh-correlation}.
This figure shows the spatial localization of the parts of the brain that BT variability was correlated with the numerical variability.
This refines the presented results in Figure~\ref{fig:unthresh-maps},
which indicated how the correlation is uniformly distributed across the brain.

  %%%%%%%%%% Corr. plot of tstats%%%%%%%%
  \begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \begin{subfigure}[ht]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/std-corr-unthresh-plot.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \end{subfigure}
      \hfill
      \begin{subfigure}[ht]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/std/correlated-unthresh.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \end{subfigure}
      \caption{Correlation of standard deviations in BT and WT from the unthresholded T-statistics.
      The first row plots the correlations voxel by voxel and is highlighted with different colors
      for two clusters, including the upper cluster (purple color) and the correlated cluster (green color).
      The second row maps the correlated area on the MNI space.}
    \label{fig:unthresh-correlation}
    \end{minipage}}
  \end{figure*}

@@

  In addition to the group-level analysis, we evaluated WT and BT variabilities between subjects individually.
  Figure~\ref{fig:unthresh-maps-sbj} shows the brain maps of standard deviations in WT and BT from unthresholded T-statistics of
  the subject with the highest WT variability among all 16 subjects.
  The average standard deviation for this subject in WT and BT was $\approx$ 0.152 and $\approx$ 0.466, respectively.
  The maps show regions in the brain where the magnitude of standard deviation exceeded 1.0 in WT in \fslafni and \afnispm,
  such as areas in the frontal and occipital lobes.
  This is similar to BT variability, even with higher variations in some voxels.

@@

We evaluated WT variability across different virtual precisions for FSL tool.
We changed the virtual precision in steps of 2 between t=1 and t=53 bits.
Figure~\ref{fig:across-precisions} represents RMSE between standard deviations of each pair of tools in BT and FSL variations across virtual precisions in WT.
%Consistant with the previous results, 
RMSE between WT and BT for \fslafni was the highest, and \fslspm was the lowest in all virtual precisions.
This also shows the similar plateaus were generated in all three pairs of tools.
Moreover, the analyses crashed for the precisions below t=11 bits during the FSL preprocessing steps. 
This needs further investigations to understand the exact failure reasons.

We determined the virtual precision of t=27 bits as the precision that minimized the RMSE between standard deviations in BT on average and WT.
This is the precision that numerical perturbations in FSL more closely simulated BT variability.
Figure~\ref{fig:gnp-mni} shows the brain maps of numerical variability obtained from the analysis with FSL at t=27 bits.
This represents regions with significant variability in the frontal, parietal, and limbic lobes.

  %%%%%%%%%% plot different precisions%%%%%%%%
  \begin{figure}[ht]
    \fbox{\begin{minipage}{\dimexpr \columnwidth-2\fboxsep-2\fboxrule}
        \centering
        \includegraphics[width=\columnwidth]{figures/rmse-precisions.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \caption{Comparison of RMSE values computed between BT and WT resulted from FSL tool for different virtual precisions.}
    \label{fig:across-precisions}
    \end{minipage}}
  \end{figure}

@@

>>>>

\section{Results}
All scripts and data to create the figures presented in this section are available at \url{https://github.com/ali4006/fuzzy-neurotools} \tristan{move to lab org}.
The computations were performed on \href{https://www.computecanada.ca}{Compute Canada's} Béluga cluster
with 872 nodes, each with 2× Intel Gold 6148 Skylake @ 2.4 GHz (40 cores/node) CPU and node memory ranging from 92 to 752 GB.

We ensured a successful replication of the analyses initally presented in~\cite{bowring2019exploring}
by visually checking the overlap \tristan{what do you mean ``visually checking the overlap?'' Afair we had looked at the difference images?} of T-statistic group maps.
We could replicate the analyses across tools with an exception for AFNI that produced results with slight differences \tristan{this needs discussion on how to present it}. 
This might be caused by the compiler variations, or different hardware and their optimizations
such as out-of-order execution (dynamic scheduling) paradigm~\cite{duben2014use,demmel2013numerical}.
In this paradigm, the processors might execute instructions out of the original order they appear based on
the availability of input data and execution units to use resources efficiently. Therefore, it might
compute floating-point operations in a different order, which often leads to different results \tristan{is this a compiler optimization? If so, just mention ``compiler optimizations such as dynamic scheduling'', no need to enter into details}.

Summary statistics for the group-level unthresholded and thresholded T-statistics \tristan{harmonize capitalization in ``T-statstics / t-statistics'' (t should be lower case unless it's the first word in the sentence)} are reported in Table~\ref{table:pipeline-stats}.
% The variability between results of each pair of tools in BT and individual tools in WT was measured using the mean and standard deviation of absolute differences.
Overall, WT variability was an order of magnitude lower than BT variability for both means and standard deviations.
While the highest variability was observed in \fslafni, \fslspm was made the least variations in BT.
Results also show that AFNI was generated the most variability in WT among the three tools \tristan{this needs statistical testing}.

%%%%%%%%%% Summary of statstics %%%%%%%%
\setlength{\tabcolsep}{7pt}
\begin{table}[h]
    \centering
    \begin{tabular}{cccc|cc}
        \toprule
        \multirow{2}{*}{}& {} & \multicolumn{2}{c}{Thresholded} & \multicolumn{2}{c}{Unthresholded} \\
        \cmidrule{3-4} \cmidrule{5-6} \\
        {} & {} & Mean & Std. dev. & Mean & Std. dev. \\
        \midrule
        \rowcolor{lightgray}
        {Between Tools} & FSL vs. SPM        &  0.043       & 1.282      & 0.242     & 0.443  \\
        \rowcolor{lightgray}
        {(BT)} & FSL vs. AFNI         &  0.099       & 1.548      & 0.302     & 0.547  \\
        \rowcolor{lightgray}
        {} & AFNI vs. SPM         &  0.079       & 1.475      & 0.254     & 0.608  \\
        {Within Tool} & FSL    &  0.005       & 0.354      & 0.034     & 0.082  \\
        {(WT)}   & SPM    &  0.006       & 0.252      & 0.025     & 0.054  \\
        {}   & AFNI   &  0.017       & 0.434      & 0.041     & 0.128  \\
        \bottomrule
    \end{tabular}
    \caption{Summary of voxel-wise mean and standard deviation of absolute differences in group-level T-statistics in BT and WT. \tristan{discuss this caption}}
    \label{table:pipeline-stats}
\end{table}

@@

%Fig 1
Comparisons of standard deviations between thresholded images \tristan{Table II should show summary statistics from the maps, is it the case?} in WT and BT
on MNI space are shown in Figure~\ref{fig:thresh-maps}.
While we observed substantial BT variations with the average standard deviation $\approx$ 1.5 \tristan{why approx? },
the magnitude of WT variations was much lower with the average standard deviation $\approx$ 0.5,
as was anticipated. However, numerical variability was still significant in WT results \tristan{this doesn't mean much. Also, ``significant'' means ``statistically significant''.}.
%Moreover, the numerical perturbations produces variations in similar order og magnitude in WT compared to BT variations
Moreover, the magnitude of WT variations reached the magnitued of BT variations
in some regions, shown in white in Figure~\ref{fig:thresh-maps}-\textbf{C}.

% Fig2
Figure~\ref{fig:dice-thresh} compares regional normalized Dice coefficients of
activated voxels, showing a linear correlation between BT and WT \tristan{report p-value}. 
Also, small P-values across all three pairs, including \num{1e-10} for \fslafni, \num{6e-04} for \fslspm,
and \num{2e-05} for \afnispm, confirmed this correlation of similarities \tristan{are these the p values of the regressions?}.
% The vertical line where the Dice score is zero in BT demonstrates regions in the brain with no overlaps between activated voxels in BT but WT,
% including left/right `V1\_ROI'\footnote{The primary visual cortex} in \fslafni and \fslspm,
% and left/right `TGd\_ROI'\footnote{The area TG dorsal (part of lateral temporal cortex)} in \afnispm results.

%%%%%%%%%% Var. of Thresh %%%%%%%%
\begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \centering
      \includegraphics[width=\textwidth]{figures/std/gl-thresh.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \caption{Thresholded group t-statistics standard deviations computed between tools (\textbf{A}) and within tools (\textbf{B}), and difference between them (\textbf{C}). }
    % so that bright blue areas indicate more similar order of magnitude of variations in both conditions,
    %and vise versa for the darkder regions.}
    \label{fig:thresh-maps}
    \end{minipage}}
  \end{figure*}

  %%%%%%%%%% Dice plot of thresholded tstats%%%%%%%%
  \begin{figure}[ht]
    \fbox{\begin{minipage}{\dimexpr \columnwidth-2\fboxsep-2\fboxrule}
    \centering
    \includegraphics[width=\columnwidth]{figures/dices_corr.png}
    \caption{Normalized, regional Dice coefficients of activated voxels for
    BT and WT using the HCP-MMP1.0 parcellation of
    ~\cite{glasser2016multi}.}
    \label{fig:dice-thresh}
    \end{minipage}}
  \end{figure}

@@

% Fig3
Similar conclusions are drawn from the unthresholded t-statistcs maps
(Figure~\ref{fig:unthresh-maps}). BT variability is an order of magnitude
higher than WT variability overall, however, BT and WT variability are
comparable in some regions.

% shows the brain maps of standard deviations for group-level unthresholded
% T-statistics in WT and BT. Results show a different order of magnitude of
% variations in WT and BT with standard deviation of $\approx$ 0.12 and
% $\approx$ 0.5 on average, respectively. We observed regions in the brain
% where the magnitude of standard deviation was close to zero in WT, but it
% exceeded 2.0 in BT, such as some parts of the limbic and frontal lobes in
% the sagittal plane. However, the brain maps included voxels with
% approximately zero differences of standard deviations in WT and BT, which
% are uniformly distributed.

%%%%%%%%%% Var. of Unthresh %%%%%%%%
\begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \centering
      \includegraphics[width=\textwidth]{figures/std/gl-unthresh.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \caption{Unthresholded group t-statistics standard deviations computed between tools (\textbf{A}) and within tools (\textbf{B}), and difference between them (\textbf{C}).}
    \label{fig:unthresh-maps}
    \end{minipage}}
  \end{figure*}

% Fig4
\tristan{rewrite this paragraph}
The scatter plot in Figure~\ref{fig:unthresh-correlation} represents the correlation of standard deviation in WT and BT variability.
We found two major clusters, including the identity cluster that corresponds to the correlations
between BT and WT with the ratio of $0.5 < BT/WT < 2$ where the standard deviations were bigger than 0.1,
and the upper cluster that shows voxels where BT $\approx$ 0.
The percentage of voxels included in the identity cluster was \%9.9 in \fslspm, \%17.3 in \fslafni, and \%13.8 in \afnispm,
and in the upper cluster was $\approx$ \%1 for each pair of tools.
The identity area is also represented on the MNI space in the second row in Figure~\ref{fig:unthresh-correlation}.
This figure shows the spatial localization of the parts of the brain that BT variability was correlated with the numerical variability.
This refines the presented results in Figure~\ref{fig:unthresh-maps},
which indicated how the correlation is uniformly distributed across the brain.

  %%%%%%%%%% Corr. plot of tstats%%%%%%%%
  \begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \begin{subfigure}[ht]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/std-corr-unthresh-plot.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \end{subfigure}
      \hfill
      \begin{subfigure}[ht]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/std/correlated-unthresh.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \end{subfigure}
      \caption{Correlation of standard deviations in BT and WT from the unthresholded T-statistics.
      The first row plots the correlations voxel by voxel and is highlighted with different colors
      for two clusters, including the upper cluster (purple color) and the correlated cluster (green color).
      The second row maps the correlated area on the MNI space. \tristan{Fix colors and captions (to be discussed)}}
    \label{fig:unthresh-correlation}
    \end{minipage}}
  \end{figure*}

@@

  Figure~\ref{fig:unthresh-maps-sbj} shows BT and WT variability of the
  unthresholded t-statistics for the subject with the highest WT
  variability among the 16 subjects. The average standard deviation for
  this subject was $\approx$ 0.152 for WT and $\approx$ 0.466 for BT
  \tristan{remove $\approx$, discuss what these numbers represent}.
  As can be seen in the Figure, WT variability approaches and even surpasses BT variability in some regions.

@@

While the previous results were obtained at the virtual precision of 53~bits for double-precision floating-point values
and 24~bits for single-precision values, we evaluated WT variability across different virtual precisions for FSL. \tristan{make sure this is described in the methods}
Figure~\ref{fig:across-precisions} represents the RMSE between standard deviations of each pair of tools in BT and FSL variations across virtual precisions in WT \tristan{clarify what is shown}.
%Consistant with the previous results, 
RMSE between WT and BT for \fslafni was the highest, and \fslspm was the lowest in all virtual precisions.
This also shows the similar plateaus were generated in all three pairs of tools.
Moreover, the analyses crashed for the precisions below t=11 bits during the FSL preprocessing steps. 
This needs further investigations to understand the exact failure reasons.

\tristan{make a full figure, similar to Fig 1, 3, and 5. }We determined the virtual precision of t=27 bits as the precision that minimized the RMSE between standard deviations in BT on average and WT.
This is the precision that numerical perturbations in FSL more closely simulated BT variability.
Figure~\ref{fig:gnp-mni} shows the brain maps of numerical variability obtained from the analysis with FSL at t=27 bits.
This represents regions with significant variability in the frontal, parietal, and limbic lobes.

  %%%%%%%%%% plot different precisions%%%%%%%%
  \begin{figure}[ht]
    \fbox{\begin{minipage}{\dimexpr \columnwidth-2\fboxsep-2\fboxrule}
        \centering
        \includegraphics[width=\columnwidth]{figures/rmse-precisions.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \caption{Comparison of RMSE values computed between BT and WT resulted from FSL tool for different virtual precisions \tristan{Make a full figure, similar to Fig 5, 3, and 1.}.}
    \label{fig:across-precisions}
    \end{minipage}}
  \end{figure}

@@

  \item[$\bullet$] Numerical instability in individual analyses is attenuated in group analyses. This is an issue for the development of fMRI biomarkers.
  \item[$\bullet$] Comparision between FSL, SPM and AFNI. Discussion on the instrumentation of math functions only.
\end{itemize}
